% This file should be replaced with your file with thesis content.
%=========================================================================
% Authors: Michal Bidlo, Bohuslav Křena, Jaroslav Dytrych, Petr Veigend and Adam Herout 2019

% For compilation piecewise (see projekt.tex), it is necessary to uncomment it and change
% \documentclass[../projekt.tex]{subfiles}
% \begin{document}

\chapter{Introduction}

TODO

\chapter{Abstract}
\label{abstrakt}

TODO 

\chapter{Theoretical Foundations}
\label{chap:theoretical-foundations}

\section{Prerequisites}

This work assumes familiarity with convolutional neural networks, gradient-based optimisation, 
and automatic differentiation as implemented in frameworks such as PyTorch~\cite{PaszkePyTorch, PyTorchDocs}. 
Goodfellow et al.~\cite{GoodfellowDeepLearning} and the official PyTorch documentation~\cite{PyTorchDocs} 
provide comprehensive introductions to these topics.

The remainder of this chapter is self-contained with respect to cellular automata and their neural extensions.

\section{Cellular Automata}

Cellular automata (CA) are discrete dynamical systems composed of a regular grid of cells, each occupying 
one state from a finite set. All cells update in parallel at each discrete time step according to a local 
rule that depends only on the states of neighbouring cells~\cite{MarinescuComplexSystems}.

A cellular automaton is defined by:
\begin{itemize}
    \item a \textbf{lattice} of cells, typically arranged in one, two, or three dimensions,
    \item a \textbf{finite state set} $\mathcal{S}$,
    \item a \textbf{neighbourhood function} $\mathcal{N}$ describing which cells influence a given cell,
    \item an \textbf{update rule} $f : \mathcal{S}^{|\mathcal{N}|} \rightarrow \mathcal{S}$ that determines the next state~\cite{IlachinskiCellular}.
\end{itemize}

Ulam and von Neumann developed the first formal CA models in the 1940s as a theoretical framework for 
self-replicating machines~\cite{VonNeumann1966}. Von Neumann's original construction used a two-dimensional 
lattice with 29 distinct cell states and demonstrated that a sufficiently complex local rule can produce 
a system capable of constructing copies of itself. This result established that self-replication does not 
require a centralised blueprint but it can emerge from purely local interactions on a discrete grid.

Wolfram later proposed a behavioural classification of CA into four qualitative classes based on their 
long-term dynamics~\cite{Wolfram1983, Wolfram2002}. Class~1 systems converge to a uniform state regardless 
of initial conditions. Class~2 systems settle into static or periodic patterns. Class~3 systems produce 
chaotic, seemingly random output sensitive to initial conditions. Class~4 systems generate complex dynamics 
with long-lived interacting structures capable of computation. This classification provides a useful 
reference for characterising the dynamics of both classical and neural CA.

\subsection{One-Dimensional Cellular Automata}

The simplest non-trivial CA operate on a one-dimensional array of cells. In the elementary CA family, 
each cell reads its own state and its two immediate neighbours (radius~1, neighbourhood size~3). With 
binary states, $2^{3} = 8$ neighbourhood configurations are possible, giving $2^{8} = 256$ distinct 
rules in Wolfram's enumeration~\cite{Wolfram1983}.

Each rule number encodes a next-state lookup table in binary. For example, Rule~90 assigns the next state 
of a cell based on the exclusive-or (XOR) of its two neighbours, producing the Sierpi\'{n}ski triangle 
from a single active seed. This pattern is fractal and self-similar, placing Rule~90 in Wolfram's Class~3 
(chaotic behaviour). Rule~30 also belongs to Class~3 and produces irregular, aperiodic evolution that has 
been used as a source of pseudorandomness~\cite{Wolfram2002}.

Rule~110 (Class~4) is the smallest known Turing-complete one-dimensional CA. Cook~\cite{Cook2004} proved 
that it supports gliders and glider collisions that can simulate any Turing machine. This result was 
significant because it showed that computational universality can arise from an extremely simple local rule 
applied to a one-dimensional lattice.

Despite their simplicity, elementary CA demonstrate all four classes of Wolfram's classification. They 
provide a minimal setting for studying how local rules produce global behaviour, a question that extends 
directly to the neural CA models discussed later in this chapter.

\subsection{Two-Dimensional Cellular Automata}

In two dimensions, the Moore neighbourhood ($3 \times 3$ block surrounding each cell) is the most common 
choice. With binary states and a Moore neighbourhood, the update rule must handle $2^9 = 512$ possible 
input configurations. Because the full lookup table is large, 2D rules are typically expressed using 
birth/survival notation. A rule written as B$n$/S$m$ specifies that a dead cell becomes alive (is ``born'') 
when it has exactly $n$ live neighbours, and a living cell survives when it has exactly $m$ live neighbours. 
All other cells die or remain dead.

Conway's Game of Life~\cite{Gardner1970} uses the rule B3/S23: a dead cell becomes alive with exactly three 
live neighbours, and a living cell survives with two or three live neighbours. Despite the simplicity of this 
rule, Life produces a wide variety of emergent structures. Still lifes (such as the block and beehive) are 
stable configurations that do not change between steps. Oscillators (such as the blinker and toad) cycle 
through a fixed sequence of states. Spaceships (such as the glider and lightweight spaceship) translate 
across the grid while cycling through their phases. More complex constructions include glider guns, which 
periodically emit spaceships, and computational structures that perform logic operations.

Berlekamp, Conway, and Guy~\cite{Berlekamp1982} showed that Life is Turing-complete: any computation that 
can be expressed as a Turing machine can, in principle, be carried out within a sufficiently large Life grid. 
This result highlights a key property of cellular automata --- complex, globally coordinated behaviour can 
emerge from a homogeneous local rule. The same property later motivated the development of neural cellular 
automata, where the local rule is learned rather than hand-designed.

\subsection{Three-Dimensional Cellular Automata}

In three dimensions, the Moore neighbourhood covers 26 adjacent cells (6 face-adjacent, 12 edge-adjacent, 
and 8 corner-adjacent) in a $3\times3\times3$ cube surrounding the central cell. Binary states with a full 
Moore neighbourhood produce $2^{27} \approx 134$~million neighbourhood configurations, making exhaustive 
rule enumeration impractical. For this reason, practical 3D~CA use totalistic or outer-totalistic rules 
that depend only on the count of live neighbours rather than their exact positions~\cite{Wolfram2002}.

Several notable 3D~CA have been studied. Von Neumann's 29-state self-replicating automaton~\cite{VonNeumann1966}, 
originally formulated in 2D, has been extended to three dimensions, where the additional spatial freedom 
allows more compact replicating structures. Excitable-media rules such as Brian's Brain have also been 
generalised to 3D lattices, producing travelling wavefronts and spiral structures in three-dimensional 
space~\cite{DeKinderen1998}. Three-dimensional Game of Life variants using totalistic birth/survival rules 
generate branching and crystal-like growth patterns that have no direct 2D analogue.

Hardware advances in the 2000s made large-scale 3D~CA simulations practical on consumer 
hardware~\cite{Tobias2009}. GPU-accelerated implementations can now run volumetric CA with grid sizes 
exceeding $128^3$ cells in real time. These volumetric models provided a natural setting for the 
neural CA extensions discussed later in this chapter, as the underlying lattice structure and 
local-update paradigm transfer directly from classical to neural formulations.

\subsection{Differentiability and Continuous Cellular Automata}

Classical CA use discrete states and discrete update rules. However, several extensions have shown that 
the core CA principles (spatial lattice, local interactions, and parallel updates) also work with 
continuous states and smooth transition functions. These continuous CA are directly relevant to neural CA, 
which operate on real-valued state vectors and use differentiable update rules trained by gradient descent.

An early theoretical connection was established by Turing~\cite{Turing1952}, whose reaction-diffusion model 
of morphogenesis describes how spatial patterns can arise from the interaction of locally diffusing chemical 
substances. Although Turing's model uses partial differential equations rather than a discrete lattice, the 
underlying principle - local interactions producing global spatial structure - is the same principle that 
governs cellular automata. This connection between reaction-diffusion systems and CA has motivated much of 
the subsequent work on continuous and neural CA.

\textbf{SmoothLife}~\cite{Rafler2011} generalises Conway's Game of Life to continuous space and time. It 
replaces the discrete neighbour count with continuous averaging kernels and the binary birth/survival rule 
with smooth sigmoid transition functions. The result is a system that runs on a continuous 2D field and 
produces gliders, oscillators, and wave-like structures that closely resemble their discrete counterparts 
but evolve smoothly in time. SmoothLife demonstrates that the characteristic behaviours of Life do not depend 
on discretisation, but rather arise from the structure of local interactions.

\textbf{Lenia}~\cite{Chan2019} extends this approach further by parameterising both the convolution kernel 
(which defines the neighbourhood shape and weighting) and the growth function (which determines how the 
local average maps to a state change). Systematic exploration of Lenia's parameter space has produced 
hundreds of stable, self-organising patterns that resemble biological microorganisms such as orbiting 
particles, pulsating organisms, and multi-species ecologies. Lenia shows that a rich space of 
emergent behaviours can be accessed by varying the parameters of a continuous local update rule.

Both SmoothLife and Lenia demonstrate that complex self-organisation does not require discrete state 
transitions. Local, differentiable interactions are sufficient to produce stable spatial structures. Neural 
cellular automata build on this observation. Instead of manually exploring the parameter space as in Lenia, 
neural CA use gradient-based optimisation to find update rules that produce a specific target pattern. The 
differentiability of the update rule is what makes this optimisation possible.

\section{Neural Cellular Automata}

Mordvintsev et al.~\cite{Mordvintsev2020} introduced Neural Cellular Automata (NCA) as a differentiable 
generalisation of classical CA. The discrete lookup-table rule is replaced by a small neural network whose 
parameters are shared across all cells. The spatial lattice stays discrete, but each cell's state becomes 
a real-valued vector, and the update rule is trained end-to-end by gradient descent.

Let the lattice be $\Omega \subset \mathbb{Z}^d$, where $d$ is the spatial dimension. Each cell 
$x \in \Omega$ holds a state vector
\[
\mathbf{s}_t(x) \in \mathbb{R}^{C},
\]
where $C$ is the number of channels. The full grid state at time $t$ is a tensor over all cell states.

The update process has two stages~\cite{Mordvintsev2020}. First, a fixed perception operator extracts 
local neighbourhood information through convolutional filters. The filters are applied to the full 
multi-channel grid and produce a perception vector at each cell location. Second, a shared neural network 
maps this perception vector to a state increment $\Delta \mathbf{s}$ that is added to the current state:
\[
\mathbf{s}_{t+1}(x) = \mathbf{s}_t(x) + \Delta \mathbf{s}(x).
\]

This residual update formulation keeps the defining CA property - strict locality - while replacing 
the hand-crafted transition function with a learnable model. Because the update rule is differentiable 
with respect to its parameters, the automaton can be unrolled over multiple time steps and trained by 
gradient descent to minimise a loss defined on the resulting configuration.

The key conceptual shift compared to classical and continuous CA is that the update rule is not designed 
or discovered by exploration, but optimised for a specific objective. The system learns to grow a target 
pattern from a seed, maintain it over time, or regenerate it after damage depending on the training 
regime. This makes neural CA a form of learned self-organisation: the local rule is shaped by optimisation, 
but the global behaviour emerges from distributed local interactions, just as in classical CA.

\section{Extension from Two to Three Dimensions}

The original NCA of Mordvintsev et al.~\cite{Mordvintsev2020} operates on a 2D grid. Extending it to 
three-dimensional voxel lattices is conceptually straightforward but introduces several practical 
differences that affect both the architecture and the training procedure.

In 2D, perception uses $3 \times 3$ filters over a nine-cell Moore neighbourhood. In 3D, the same 
filters generalise to $3 \times 3 \times 3$ volumes covering twenty-seven cells. Prior 3D 
extensions~\cite{Sudhakaran2021,Rossi2021,Aadityaza2024} have shown that Sobel-type gradient filters 
and other perception operators transfer directly to $x$, $y$, and $z$ axes without changing the 
overall update formulation.

However, volumetric grids introduce two important differences. First, cell count grows cubically with 
resolution. A $32^3$ grid contains 32\,768 cells; a $64^3$ grid contains 262\,144. This cubic scaling 
increases both memory consumption and computational cost per step, particularly during training when 
gradients must be stored for backpropagation through multiple steps.

Second, information still propagates only through local interactions, one step at a time. In a $32^3$ 
grid, a signal originating at one corner must traverse roughly 55 steps (the grid diagonal in Manhattan 
distance through the $3 \times 3 \times 3$ neighbourhood) to reach the opposite corner. In larger grids, 
this propagation distance grows further. Maintaining coherent global structure therefore requires stable 
dynamics over longer unrolling horizons than in comparable 2D systems.

As a consequence, behaviours such as persistence and regeneration may 
behave differently in volumetric domains. The change in neighbourhood structure (27 neighbours instead of 
9), spatial topology, and information propagation distance can all affect the stability and character of 
the emergent dynamics.

\section{Core Components of a 3D Neural Cellular Automaton}
\label{sec:core-components}

This section describes the principal components of a three-dimensional neural cellular automaton at the 
conceptual level. The goal is to establish the general design principles that underlie these systems before 
addressing the specific architecture and implementation in later chapters.

\subsection{State Representation}

Each voxel in a 3D neural CA maintains a vector of $C$ real-valued channels:
\[
    \mathbf{s} \in \mathbb{R}^{C}.
\]

The channels divide into two functional groups. \textbf{Visible channels} represent quantities 
compared directly to a target pattern, such as colour components (RGB), occupancy, or material 
properties. \textbf{Hidden channels} provide internal memory that is not directly supervised but 
carries information between neighbouring cells across time steps~\cite{Mordvintsev2020}.

The separation between visible and hidden channels is a design choice inherited from the 2D NCA 
formulation. Visible channels define the observable output of the system, while hidden channels 
give each cell additional representational capacity to encode information about its local context, 
history, or role within the emerging structure. Because updates are strictly local, a cell cannot 
directly observe distant parts of the grid. Hidden channels allow cells to accumulate and relay 
information over multiple steps, enabling the kind of long-range coordination needed to produce 
coherent global patterns from local rules.

Most NCA models also designate one channel as a continuous \emph{alive} signal. This channel 
determines whether a cell is considered active and whether its updates are applied. The alive 
mechanism appears in the original NCA~\cite{Mordvintsev2020} and has parallels in continuous CA 
such as SmoothLife~\cite{Rafler2011} and Lenia~\cite{Chan2019}, where real-valued fields govern 
the activation and deactivation of spatial regions.

\subsection{Perception Mechanism}

Neural CA preserve the locality constraint that defines classical CA: each cell updates based only 
on its immediate neighbourhood~\cite{IlachinskiCellular, Wolfram2002}. In a 3D neural CA, locality 
is realised by applying a fixed set of convolutional filters with kernel size $3 \times 3 \times 3$ 
to the multi-channel state grid. Each filter reads the 27-cell neighbourhood of every voxel and 
produces a scalar output per channel~\cite{Mordvintsev2020, Sudhakaran2021}.

Common filter types include:
\begin{itemize}
    \item \textbf{Identity filters}, which return the cell's own state unchanged. These provide the 
    update rule with information about what the cell currently contains.
    \item \textbf{Neighbour aggregation filters}, which compute a sum or average of adjacent voxels. 
    These capture the local density and composition of the neighbourhood.
    \item \textbf{Gradient filters}, such as Sobel-type kernels extended to the $x$, $y$, and $z$ axes. 
    These approximate directional differences and provide a local analogue of gradient sensing 
    in biological morphogenesis~\cite{Mordvintsev2020, Sudhakaran2021, Turing1952}.
\end{itemize}

The outputs of all filters applied across all channels are concatenated into a single perception vector 
per cell. If the state has $C$ channels and $G$ filter groups are used, the perception vector has 
dimensionality $G \cdot C$. This vector is then passed to the update function as input.

The filters are fixed and non-learnable. This is a deliberate design choice: using fixed filters 
constrains the form of spatial information available to the update rule, which stabilises training 
and reduces the number of parameters that must be optimised. The approach follows the convention 
established by Mordvintsev et al.~\cite{Mordvintsev2020} and adopted in subsequent 
work~\cite{Sudhakaran2021, Rossi2021, Aadityaza2024}.

\subsection{Update Rule}

The update rule is a shared parametric function that governs how each cell's state evolves over time. 
It takes the perception vector as input and produces a state increment $\Delta \mathbf{s}$ added to 
the current state:
\[
\mathbf{s}_{t+1} = \mathbf{s}_t + \Delta \mathbf{s}.
\]

This residual formulation encourages incremental modifications rather than abrupt state changes, 
contributing to dynamical stability over long simulation horizons~\cite{Mordvintsev2020}.

The update function is typically implemented as a small multilayer perceptron (MLP) or a sequence of 
$1 \times 1 \times 1$ convolutions applied independently at each cell. Point-wise application ensures 
that the update depends only on the locally perceived information, not on any global state. The 
parameter count is kept deliberately low following the original 2D model, approximately 8\,000 trainable 
parameters~\cite{Mordvintsev2020}. This small scale underlines that the complexity of the resulting 
structures arises from iterative local interaction, not from network capacity.

Two stabilisation mechanisms appear in most NCA formulations:
\begin{itemize}
    \item \textbf{Alive gating} conditions updates on the alive channel. If a cell and all of its 
    neighbours have alive values below a threshold, the cell's update is set to zero. This prevents 
    empty regions from spontaneously generating structure and provides a soft boundary between active 
    and inactive areas of the grid~\cite{Mordvintsev2020}.
    \item \textbf{Stochastic update masking} applies updates to a random subset of cells at each step 
    rather than to all cells simultaneously. This breaks temporal synchronisation artefacts and 
    encourages robustness to asynchronous perturbations~\cite{Mordvintsev2020}.
\end{itemize}

Both mechanisms improve long-term stability and regenerative capacity. They become increasingly 
important as systems scale to larger grids or higher dimensions, where dynamical instabilities 
can accumulate over more cells and longer simulation horizons.

\subsection{Grid Management}

The neural CA runs on a finite three-dimensional lattice with fixed dimensions. Boundary conditions 
(periodic, zero-padded, or reflective) are chosen per experiment and affect the dynamics near the 
grid edges~\cite{Mordvintsev2020}. Sudhakaran et al.~\cite{Sudhakaran2021} note that the choice of 
boundary handling varies across implementations and can influence the shapes that the automaton produces.

The primary challenge of 3D grids is their cubic memory scaling. A grid of size $N^3$ with $C$ channels 
per cell and 32-bit floating-point precision requires $N^3 \times C \times 4$ bytes of storage per state. 
For a $32^3$ grid with 20 channels, this is approximately 2.5~MB per state. During training, multiple 
states are held simultaneously (for batching and pool-based sampling), and gradient storage can multiply 
memory requirements further.

Research on large-scale volumetric simulations addresses this through specialised data structures. 
Sparse Voxel Octrees~\cite{Laine2010} exploit the observation that most voxels in a large grid are 
empty, storing only occupied regions in a hierarchical tree. OpenVDB~\cite{Museth2013} provides a 
similar approach with dynamic topology, allowing efficient insertion and deletion of voxels during 
simulation. While the framework developed in this work uses dense grids at moderate resolution, these 
sparse representations offer a path toward scaling neural CA to larger volumetric domains in future work.

Initialisation is another practical consideration. Most neural CA start from a \emph{seed} - a single 
active cell or small cluster placed at the centre of an otherwise empty grid. The alive channel of the 
seed cell is set to a positive value, and the system must grow the target structure from this minimal 
starting point through repeated application of the local update rule.

\subsection{Training and Optimisation}

Because the update rule is differentiable, the parameters of a neural CA can be optimised by gradient 
descent. The standard training procedure unrolls the automaton for $T$ steps starting from a seed 
state, computes a loss between the resulting configuration and a target, and backpropagates through 
the full unrolled computation graph~\cite{Mordvintsev2020}.

The most common loss function is mean squared error (MSE) between the predicted and target visible 
channels. For structural objectives, intersection-over-union or other task-specific 
metrics may be used~\cite{Niklasson2021}. The optimiser is typically Adam or one of its 
variants, with learning rates in the range of $10^{-4}$ to $10^{-2}$~\cite{Sudhakaran2021}.

Several training strategies have been developed to encourage specific behaviours:
\begin{itemize}
    \item \textbf{Pool-based sampling} maintains a pool of persistent states at various stages of growth. 
    Each training iteration samples a batch from the pool, runs the automaton forward, computes the loss, 
    and writes the updated states back. This exposes the model to states at different developmental stages 
    rather than always starting from a fresh seed~\cite{Mordvintsev2020}.
    \item \textbf{Curriculum learning} gradually increases the number of unrolled steps during training. 
    Early iterations use few steps to learn coarse shape, while later iterations use many steps to refine 
    details and test long-term stability.
    \item \textbf{Training regime selection} determines which aspects of the dynamics are explicitly 
    optimised. These regimes (growth, persistence, regeneration) are discussed in detail in 
    the Related Work chapter.
\end{itemize}

Backpropagation through many unrolled steps requires storing intermediate activations for the entire 
sequence, which can exhaust GPU memory. Gradient checkpointing~\cite{PaszkePyTorch} addresses this by 
discarding intermediate activations during the forward pass and recomputing them during the backward 
pass, trading approximately twice the computation for substantially reduced memory usage.

%=========================================================================
\chapter{Related Work}

This chapter reviews the research most directly related to neural cellular automata and their 
three-dimensional extensions. The focus is on architectures, training procedures, and experimental 
results. Each section identifies limitations or open questions that motivate the framework developed 
in this work.

Several of the works discussed here were already introduced in Chapter~\ref{chap:theoretical-foundations} 
to establish formal background. Here, the focus shifts to their concrete design decisions, findings, and the 
specific choices that inform the approach taken in this work.

\section{Growing Neural Cellular Automata}

Mordvintsev, Randazzo, Niklasson, and Levin~\cite{Mordvintsev2020} introduced the Growing Neural 
Cellular Automata paradigm that serves as the methodological foundation for this work. They replaced 
the discrete lookup-table rules of classical CA with a small shared neural network trained end-to-end 
by gradient descent. The model preserves strict spatial locality while learning to grow and maintain 
target patterns from simple initial conditions - typically a single live cell.

Unlike earlier continuous CA such as Lenia~\cite{Chan2019}, which rely on manual parameter exploration 
to discover interesting behaviours, neural CA use supervised optimisation. The transition function is 
trained to produce a specific target pattern. This shift from rule discovery to task-driven optimisation 
defines the approach extended to three dimensions in this work.

Chapter~\ref{chap:theoretical-foundations} covered the general NCA principles. The focus here is on the 
concrete design and empirical properties of the 2020 implementation: its architecture, stabilisation 
mechanisms, training procedure, and experimental results.

\subsection{State Representation}

Each cell in the 2D model holds a 16-channel real-valued state vector. Four channels encode RGBA colour 
values compared against a target image. The remaining twelve channels serve as hidden memory.

The hidden channels function as a distributed memory system embedded in the lattice. Because updates are 
strictly local, long-range coordination must emerge through repeated short-range interactions. A cell 
cannot observe the state of a distant cell directly, it can only receive information that has been 
relayed through intermediate cells over multiple steps. The hidden channels provide the representational 
capacity for this relay process.

This design is central to the stability and regenerative behaviour observed in later experiments. 
Reducing the number of hidden channels degrades pattern quality and stability, while increasing it 
beyond a certain point yields diminishing returns~\cite{Mordvintsev2020}. The balance between having 
enough memory for coordination and keeping the state compact enough for efficient training directly 
informs the channel configurations used in subsequent 3D extensions.

\subsection{Perception Mechanism}

Perception uses a fixed set of non-learnable convolutional filters applied to the full multi-channel grid:

\begin{itemize}
    \item An identity filter, returning each cell's own state,
    \item Sobel filters approximating spatial derivatives in the $x$ and $y$ directions,
    \item A Laplacian filter capturing local curvature.
\end{itemize}

Applied across all 16 channels, these three filter types produce a $3 \times 16 = 48$-dimensional 
perception vector at each spatial location. The Sobel filters detect directional gradients in the 
neighbourhood, for example where a pattern transitions from occupied to empty, while the Laplacian 
identifies edges and boundaries by measuring how much a cell's state differs from its neighbours' average.

A small MLP operating independently at each cell transforms this perception vector into a state increment. 
The perception step does not introduce any learnable parameters; all learning occurs in the MLP.

Fixed perception filters are used to ensure locality and translation invariance by design, while also stabilising 
optimisation by limiting the spatial information available to the update rule. This choice becomes particularly 
important in volumetric extensions, where the neighbourhood expands from 9 to 27 cells and the perception 
vector grows correspondingly in dimensionality.

\subsection{Update Rule and Parameter Sharing}

The update rule is a lightweight neural network shared across all cells. It produces a state 
increment added residually to the current state:
\[
\mathbf{s}_{t+1} = \mathbf{s}_t + \Delta \mathbf{s}.
\]

Residual updates encourage incremental modifications rather than abrupt transitions, contributing 
to dynamical stability. Because the same parameters apply at every spatial location, the learned 
rule defines a homogeneous dynamical system over the entire grid. No cell has a special role by 
construction, this means that any differentiation in cell behaviour must emerge from the spatial 
context encoded in the perception vector.

The full architecture has approximately 8\,000 trainable parameters. This small scale is deliberate: 
it shows that the complexity of the resulting structures comes from iterative local interaction, not 
from model capacity. By contrast, a standard convolutional neural network for image generation may 
have millions of parameters. The NCA achieves comparable visual complexity through temporal depth 
(running many update steps) rather than through network width.

\subsubsection{Stabilisation Mechanisms}

Two mechanisms are critical for robust long-term dynamics:

\begin{itemize}
    \item \textbf{Alive gating} conditions state updates on whether the designated alpha channel 
    exceeds a threshold in the local neighbourhood. This creates a soft alive/dead boundary and 
    prevents inactive regions from destabilising active structures. Without alive gating, empty 
    regions can spontaneously develop non-zero states due to numerical noise, leading to spurious 
    growth outside the intended pattern.
    \item \textbf{Stochastic update masking} applies updates to a random subset of cells at each 
    step. At each time step, each cell independently decides (with some probability, e.g., 0.5) 
    whether to apply its computed update or skip it. This breaks synchronisation artefacts - patterns 
    that form only because all cells update in lockstep - and produces dynamics that are robust to 
    asynchronous perturbations.
\end{itemize}

Both mechanisms improve long-term stability and regenerative capacity. Horvath and Levin~\cite{Horvath2021} 
later demonstrated that stochastic masking also encourages the emergence of distinct cell types within the 
same NCA, as cells must develop local strategies that work regardless of which neighbours updated in the 
current step.

\subsection{Training Regimes}

Mordvintsev et al.~\cite{Mordvintsev2020} introduced three training regimes: growing, persistent, 
and regenerating. Each regime targets a different aspect of the learned dynamics. The choice of regime 
determines the balance between initial pattern formation, long-term stability, and robustness to damage. 
These regimes define the training methodology adopted and extended in this work.

\subsubsection{Growing Regime}

The model is trained to produce a target image from a single active seed. Loss is computed after 
a fixed number of update steps $T$, and the model learns to expand outward and approximate the 
target pattern within that horizon.

No constraint is imposed beyond step $T$. Patterns may drift, decay, or diverge if simulation 
continues past the trained step count. In practice, growing-only models commonly exhibit instability 
within 100-200 steps beyond their training horizon~\cite{Mordvintsev2020}.

\subsubsection{Persistent Regime}

The persistent regime evaluates loss at multiple time steps beyond initial convergence. Rather than 
stopping at step $T$, training continues to evaluate the model at steps $T$, $T + \delta$, $T + 2\delta$, 
and so on, penalising any drift away from the target.

This produces more stable dynamics. Mordvintsev et al. observed that persistent models also exhibit 
limited regenerative behaviour even without explicit damage training, suggesting that the stability 
mechanisms learned for persistence partially generalise to perturbation recovery.

Randazzo et al.~\cite{Randazzo2021} later extended this idea by training NCA with persistent memory, 
showing that models trained for stability over very long horizons develop increasingly robust internal 
representations in their hidden channels.

\subsubsection{Regenerating Regime}

During training, random damage is applied by zeroing out contiguous rectangular regions of the grid. 
The model must reconstruct the target from the damaged configuration.

This explicit objective produces the most robust dynamics. Regenerating models recover from substantial 
perturbations and maintain structural integrity over long simulation horizons. The regenerative capacity 
is qualitatively reminiscent of biological organisms that can regrow damaged tissue from partial 
structure~\cite{Mordvintsev2020}.

\subsection{Relevance to This Work}

This work adopts the architecture, training regimes, and optimisation strategy of Mordvintsev 
et al.~\cite{Mordvintsev2020} as its baseline and extends the framework to three-dimensional voxel 
lattices. The transition from 2D to 3D increases neighbourhood size from 9 to 27 cells, increases 
computational cost cubically, and extends the information propagation distance required for global 
coordination. Whether the stability and regenerative behaviours observed in planar domains carry over 
to volumetric grids is a central question motivating the experiments in this work.

\section{Three-Dimensional Neural Cellular Automata in Interactive Environments}

Sudhakaran et al.~\cite{Sudhakaran2021} extended NCA to three dimensions within the sandbox game 
\emph{Minecraft}, evaluating volumetric growth of structures and functional machines under real-time 
constraints and user interaction.

Their model extends the 2D growing NCA paradigm to a 3D voxel grid. Each voxel stores a state 
vector with visible and hidden channels. Perception uses local $3 \times 3 \times 3$ neighbourhoods 
with identity and gradient features, processed by a small neural network applied identically at every 
voxel. The system is trained by gradient descent to grow a target structure from a single seed.

The authors demonstrated that neural CA can produce coherent volumetric structures in a dynamic, 
interactive environment. The learned rules stay local and homogeneous, preserving the CA paradigm while 
enabling complex global structure formation. Their experiments showed convergence toward target voxel 
configurations and partial regenerative behaviour under perturbation - when parts of a Minecraft 
structure were manually destroyed, the NCA could partially rebuild the damaged region.

A notable contribution of this work is its evaluation in a constrained real-time environment. The 
Minecraft platform imposes fixed block types, a specific rendering pipeline, and interaction mechanisms 
that differ from the unconstrained continuous-valued grids used in purely computational experiments. 
Despite these constraints, the NCA paradigm produced recognisable structures, demonstrating that the 
approach is not limited to idealised simulation environments.

\subsection{Experimental Scope and Constraints}

The evaluation focuses on proof-of-concept demonstrations: successful growth toward predefined targets 
and qualitative assessment of structural coherence. Quantitative metrics for long-term stability, 
systematic regeneration under controlled damage patterns, or comparison with other 3D generative 
methods are limited.

The \emph{Minecraft} environment also constrains the experimental setup. Voxel states are mapped to 
a discrete set of block types, the rendering pipeline is fixed, and interaction occurs through the 
game's built-in tools. While this provides a real-time testbed accessible to a broad audience, it 
limits geometric complexity, material expressiveness, and the precision of experimental measurements 
available for systematic investigation.

\subsection{Relation to This Work}

Sudhakaran et al.~\cite{Sudhakaran2021} provide the methodological basis for extending neural CA 
from planar grids to volumetric $3 \times 3 \times 3$ perception neighbourhoods. The preservation 
of strict locality, parameter sharing, and gradient-based optimisation forms the conceptual foundation 
adopted here.

The framework developed in this work embeds the 3D neural CA within Blender instead of a game engine. 
Blender provides programmatic access to volumetric grids, mesh representations, rendering pipelines, 
and Python scripting through its built-in API. This gives precise control over lattice resolution, 
boundary conditions, material properties, and intermediate-state inspection.

The goal is not only to demonstrate 3D feasibility, but to provide a research-oriented framework 
for controlled investigation of stability, persistence, and regeneration in volumetric neural CA. 
Where the Minecraft work showed that 3D NCA can produce visually convincing results in a game 
environment, this work aims to provide the tools for systematic experimentation and analysis.

\section{Open-Source and Applied Three-Dimensional Neural Cellular Automata}

Beyond the seminal work of Mordvintsev et al.~\cite{Mordvintsev2020} and the interactive extension 
by Sudhakaran et al.~\cite{Sudhakaran2021}, several open-source projects have explored 
three-dimensional neural CA. This section reviews implementations whose design decisions are 
directly relevant to the framework developed in this work.

\subsection{Standard 3D Reconstruction Baseline}

The repository of Aadityaza~\cite{Aadityaza2024} implements a direct extension of the Growing NCA 
paradigm to three-dimensional space. It uses a Jupyter Notebook environment and PyTorch to 
reconstruct voxel shapes from a single seed.

This project provides a clean verification of 3D perception operators, specifically the expansion 
of Sobel-type filters into tri-axial gradient kernels for the $x$, $y$, and $z$ axes. The 
implementation confirms that the convolution-based perception approach used in 2D NCA applies 
directly to volumetric tensors without losing spatial locality. For this work, the project served 
as an early reference point for validating the feasibility of the 3D extension.

The implementation is focused on single-target reconstruction and does not include a training server, 
interactive visualisation, or modular architecture for experimenting with different configurations. 
These are gaps that the framework developed in this work addresses.

\subsection{Voxel-Based Morphogenetic Experiments}

Rossi and collaborators~\cite{Rossi2021} explored voxel-based neural CA for three-dimensional 
morphogenesis. Their implementation extends the 2D growing NCA to volumetric grids with 
$3 \times 3 \times 3$ perception filters and hidden channel representations.

The system trains to approximate 3D voxel targets from simple seeds. Its architecture follows 
the 2D formulation closely: fixed perception filters, residual updates, shared parameters, and 
gradient-based optimisation. The experiments confirm that volumetric growth works under strict 
locality constraints. However, the evaluation is limited to demonstrating successful growth toward 
targets. Systematic analysis of long-term persistence, regeneration capacity, or scaling behaviour 
under increasing resolution is not provided.

Both open-source projects provide proof-of-concept validation for 3D morphogenetic NCA but leave 
open questions about stability and robustness in larger volumetric domains. They also lack the 
integration with a 3D modelling environment that would allow researchers to interactively visualise 
and manipulate the NCA during training.

\section{Chapter Conclusion}

The works reviewed in this chapter established the architectural principles such as local perception, 
shared neural update rules, residual dynamics, and gradient-based optimisation, that form the 
methodological basis for this work. The original 2D Growing NCA demonstrated that a small neural 
network with fewer than 10\,000 parameters can learn to grow, maintain, and regenerate complex 
patterns through purely local interactions. Subsequent three-dimensional extensions confirmed that 
these principles transfer to volumetric grids and interactive environments, though systematic 
evaluation in 3D remains limited.

The framework developed in this work builds on these foundations. It integrates volumetric 
perception, shared update networks, and structured training regimes within a controlled 3D 
setting. Unlike prior implementations, it is designed to be modular and extensible, providing a 
general experimental environment where different architectures, training strategies, and evaluation 
procedures can be compared systematically. The integration with Blender provides interactive 
visualisation and mesh-based target specification, addressing the tooling limitations identified 
in existing open-source 3D NCA projects.

\chapter{Architectural Design}

The following chapters present the design of the 3D NCA framework and its overall architecture. 
They introduce the key components that form the framework and describe how these components interact 
with each other. The aim of this section is to provide a clear conceptual overview of the system and 
explain the roles of its individual parts within the overall structure. To maintain focus on the 
architectural design, implementation-specific details like libraries and frameworks chosen are intentionally 
omitted at this stage. These details are discussed separately in later sections 
dedicated to the implementation of the framework.

\section{Design Goals and Requirements}

The purpose of the framework is to provide an environment for experimenting with neural cellular automata 
operating on three-dimensional voxel grids. In order to support experimentation and future extensions, the 
architecture must provide a clear structure and well-defined interactions between its components.

The primary design goal of the framework is to enable simulation and training of 3D neural cellular automata 
while maintaining a modular and extensible architecture. The system should allow individual components, such 
as the simulation grid, update mechanism, and training pipeline, to be developed and modified independently. 
This approach simplifies experimentation with different models and update rules.

Another important design objective was to keep the core functionality of the framework independent from 
visualization tools and external applications. The framework was therefore designed as a standalone module 
that provides the fundamental functionality required for initializing, adjusting and training neural cellular automata. 
Additional systems can build on top of this module while reusing the core computational components.

During the development process, practical constraints related to visualization and execution environments 
influenced the architectural design. In particular, combining interactive visualization with computationally 
intensive training within a single environment proved difficult. These limitations motivated a separation 
between the components responsible for computation and those responsible for visualization and user 
interaction.

As a result, the system evolved towards an architecture where the core framework provides the underlying 
simulation and training functionality, while additional components utilize this functionality to provide 
visualization and interaction capabilities. This approach preserves the independence of the framework while 
allowing it to be integrated into different workflows and environments.

\section{System Overview}
TOOD

\section{Core Components of the Framework}
TODO

\chapter{Implementation}
TODO

\chapter{Testing}
TODO

\chapter{Experiments}
TODO

\chapter{Conclusion}
\label{zaver}

TODO

%=========================================================================

% For compilation piecewise (see projekt.tex), it is necessary to uncomment it
% \end{document}